Large squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-07-10 large_number_squeezing_with_atoms.tex
Fixing minor typos.

\begin{abstract}
We examine the feasibility of creating and measuring large relative number squeezing in multicomponent trapped Bose-Einstein condensates.  In the absence of multimode effects, this squeezing can be arbitrarily large, but a range of processes limit the measurable squeezing in realistic trap configurations.  We examine these processes, and suggest methods to mitigate them. We conclude that high levels of squeezing with large numbers of atoms is feasible, but can realistically only be achieved in traps with particular trap geometries.  We also introduce a method of maximising the measurable squeezing by using a pi-pulse during the process to improve spatial mode-matching.
\end{abstract}

@@

In section \ref{secTwoModeAnalytic} we examine the ideal behaviour when the system is reduced to a single spatial mode, extracting a full analytic solution to the squeezing for arbitrary \textbf{whatever it is we extended, along with reference of what was covered previously.}  In section \ref{sec:MMdescription} we explore the idea of how multimode behaviour will become important in physical systems, and introduce our model for simulating higher dimensional trap geometries.  Sec.~\ref{sec:1Dsim} analyses one-dimensional models, highlighting some of the issues regarding optimising squeezing parameters and mode-matching in squeezing experiments.  Sec.~ref{sec:pipulse} shows that the application of a $\pi$-pulse during the process can dramatically improve mode-matching issues.  Sec.~\ref{sec:2D3Dsqueezing} investigates measurable squeezing in 2D and 3D, and uses a Bogoliubov analysis to explain the  scaling with dimensionality.  Sec.~\ref{sectionConclusion} describes the conclusion that large volume, flat-bottomed, three-dimensional traps, which are obviously compatible with large atom number, can produce extremely high degrees of squeezing.

@@

We begin by considering a simplified version of the problem, assuming that only one of these spatial modes is relevant for each of the condensate's components.  This is done by writing the multimode atomic field operators in terms of a set of basis functions $\hat{\psi}_a({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{a}_{n}(t)$ and $\hat{\psi}_b({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{b}_{n}(t)$, where $\hat{a}_{n}$ and $\hat{b}_{n}$ annihilate a particle with normalised spatial mode $u_n({\mathbf{r}})$ and internal state $|a\rangle$ or $|b\rangle$ respectively. With this expansion, the nonlinear term in the multimode Hamiltonian for the self-interaction of the $|a\rangle$ internal state becomes
\begin{equation}
\hat{H} = \frac{U_{aa}}{2} \int dV \sum_{nmpq} u_n^* u_m^* u_p u_q \, \hat{a}^{\dagger}_{n} \hat{a}^{\dagger}_{m} \hat{a}_{p} \hat{a}_{q} 
\end{equation}
When the system remains predominantly in one mode, we can approximate this sum as the single term $\hat{H} = \hbar\chi_{aa} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}/2$ where
\begin{equation}
\chi_{aa} = \frac{1}{\hbar} U_{aa} \int |u_{i}({\mathbf{r}})|^4 \, dV.
\label{eqChiUequivalence}
\end{equation}
We proceed similarly for the other terms.  In the parameter regimes we are interested in, the detuning between the two spatial modes will be irrelevant, so setting our zero of energy appropriately, we are left with only a standard Kerr-type nonlinearity arising from atom-atom interactions, as well as a linear coupling between the two modes.  The Hamiltonian governing this reduced system is given by
\begin{equation}
\hat{H} = \hbar\frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \hbar\chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \hbar\frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}
          + \hbar\Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger}  \hat{a} )
\label{eqTwoModeHamiltonian}
\end{equation}
where the $\chi_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\Omega$ couples the two modes and allows for population transfer between them.

The system is prepared with all the population in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to be occupied as a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number.  As our simulation and experiment are insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, or a mixture of Poissonian-distributed number states.  This state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

@@

In Figure \ref{figTwoModeAnalyticExamples} we plot a solution for the normalized number variance in mode $|a\rangle$ as a function of $\phi$, choosing a typical set of nonlinearities. This example illustrates the basic features of the system: During the final coupling period, the normalised number variance undergoes cyclic variation with a period $\tau=\pi / \Omega$, and has a series of minima. It is necessary to choose both the correct phase $\phi$ as well as the correct length of time for the coupling to be applied in order to obtain optimum squeezing.

@@

The first problem is present even if the spatial modes are uncoupled.  The two-mode model shows that the parameters required to obtain best squeezing are dependent on the strength of the nonlinearity. In the multimode case, the effective nonlinearity of a mode is a function of the mode shape, as is shown by Eq.~(\ref{eqChiUequivalence}). The optimum hold time $\tau_{\mathrm{hold}}$ depends on the product of the effective mode nonlinearity and number number of particles in that mode, as does the recombination phase $\phi$.  Different modes in a multimode environment will typically have varying effective nonlinearity-particle number products and therefore squeeze at different rates, with each mode having a different optimum $\tau_{\mathrm{hold}}$ and optimum recombination phase $\phi$. This means that the best squeezing will be achieved by choosing values of $\tau_{\mathrm{hold}}$ and $\phi$ that, averaged across all the spatial modes present, result in the overall lowest total number variance. Clearly, however, this averaged number variance will be still be higher than the specific mode that exhibits the best squeezing.

@@

An analytic solution to the multimode, higher-dimensional problem is intractable, so we must use a numerical solution.  We use stochastic methods based on phase space representations to simulate the dynamics of the multimode quantum fields.  These methods alleviate the problem of directly simulating elements of the Hilbert space, the size of which increases exponentially with the number of spatial modes \cite{gardiner1991, steelET1998}.  Stochastic methods achieve this by finding a sufficiently well-behaved quasi-probability representation for the density matrix, which can then be simulated as the average behaviour of a number of low-dimensional samples.  These sample objects are the same size as the equivalent classical field.  In our case, the only tractable phase space representation is the functional Wigner representation, which is well behaved when the quantum field is approximately Gaussian, as it is for coherent and squeezed states.  Stochastic methods based on the functional Wigner representation have been used to model the behaviour of BECs and atom lasers \cite{johnssonET2007,dallET2009,dennisET2010}, as the Wigner representation allows the mapping of the system to a Fokker-Planck equation with positive semi-definite diffusion matrix (unlike the $P$-representation), and does not suffer exponential path-weighting problems that would constrain it to very short simulation intervals (as would the positive-$P$ representation).

@@

The simulation shown in Fig.~\ref{fig:3DBogScaling} verified that the number of accessible states for three-dimensional traps decreases with volume in 3D.  This is because the energy of the ground state of the BEC is reducing faster than the density of states at low energies is increasing.  This scaling of the estimated Bogoliubov fraction in 3D suggests that strong squeezing may actually be found in very large, weakly trapped condensates, provided the trapping potentials are box-like.  Although systematic analysis of squeezing across a wide range of large 3D traps and interaction strengths is impossible due to the breakdown of the TWA method, fortunately the method is accurate when the highest squeezing is produced.  While inter-state scattering lengths of most atomic species and states are not well characterised, extremely high squeezing can be found in systems with large, three-dimensional box-like potentials, and asymmetry in the three scattering lengths.  A sample 3D simulation with this geometry \textbf{with some lengths and some atom number}, and only a single non-zero scattering length, gives a result of 15dB of squeezing.  This system produces an extremely low non-condensed fraction of $0.0004$, in agreement with the Bogoliubov calculations.  
%From the "winner winner chicken dinner" email:  Simulating a 3D BEC in a box with the same parameters as the experimental paper works well. Bog mode fraction is around 4e-4, and we get squeezing of around 0.026. This is assuming we can set two scattering lengths to zero, and have the third equal to that of Rb87.

>>>>

\begin{abstract}
We examine the feasibility of creating and measuring large relative number squeezing in multicomponent trapped Bose-Einstein condensates.  In the absence of multimode effects, this squeezing can be arbitrarily large for arbitrarily large condensates, but a range of processes limit the measurable squeezing in realistic trap configurations.  We examine these processes, and suggest methods to mitigate them. We conclude that high levels of squeezing with large numbers of atoms is feasible, but can realistically only be achieved in traps with particular trap geometries.  We also introduce a method of maximising the measurable squeezing by using a pi-pulse during the process to improve spatial mode-matching.
\end{abstract}

@@

In section \ref{secTwoModeAnalytic} we examine the ideal behaviour when the system is reduced to a single spatial mode, extracting a full analytic solution to the squeezing for arbitrary \textbf{whatever it is we extended, along with reference of what was covered previously.}  In section \ref{sec:MMdescription} we explore the idea of how multimode behaviour will become important in physical systems, and introduce our model for simulating higher dimensional trap geometries.  Section~\ref{sec:1Dsim} analyses one-dimensional models, highlighting some of the issues regarding optimising squeezing parameters and mode-matching in squeezing experiments.  Section~\ref{sec:pipulse} shows that the application of a $\pi$-pulse during the process can dramatically improve mode-matching issues.  Section~\ref{sec:2D3Dsqueezing} investigates measurable squeezing in 2D and 3D, and uses a Bogoliubov analysis to explain the  scaling with dimensionality.  Section~\ref{sectionConclusion} describes the conclusion that large volume, flat-bottomed, three-dimensional traps, which are obviously compatible with large atom number, can produce extremely high degrees of squeezing.

@@

We begin by considering a simplified version of the problem, assuming that only one of these spatial modes is relevant for each of the condensate's components.  This is done by writing the multimode atomic field operators in terms of a set of basis functions $\hat{\psi}_a({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{a}_{n}(t)$ and $\hat{\psi}_b({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{b}_{n}(t)$, where $\hat{a}_{n}$ and $\hat{b}_{n}$ annihilate a particle with normalised spatial mode $u_n({\mathbf{r}})$ and internal state $|a\rangle$ or $|b\rangle$ respectively. With this expansion, the nonlinear term in the multimode Hamiltonian for the self-interaction of the $|a\rangle$ internal state becomes
\begin{equation}
\hat{H} = \frac{U_{aa}}{2} \int dV \sum_{nmpq} u_n^* u_m^* u_p u_q \, \hat{a}^{\dagger}_{n} \hat{a}^{\dagger}_{m} \hat{a}_{p} \hat{a}_{q} 
\end{equation}
When the system remains predominantly in spatial mode $u_0(\mathbf{r})$, we can approximate this sum as the single term $\hat{H} = \hbar\chi_{aa} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}/2$ where
\begin{equation}
\chi_{aa} = \frac{1}{\hbar} U_{aa} \int |u_{0}({\mathbf{r}})|^4 \, dV.
\label{eqChiUequivalence}
\end{equation}
We proceed similarly for the other terms.  In the parameter regimes we are interested in, the detuning between the two spatial modes will be irrelevant, so setting our zero of energy appropriately, we are left with only a standard Kerr-type nonlinearity arising from atom-atom interactions, as well as a linear coupling between the two modes.  The Hamiltonian governing this reduced system is given by
\begin{equation}
\hat{H} = \hbar\frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \hbar\chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \hbar\frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}
          + \hbar\Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger}  \hat{a} )
\label{eqTwoModeHamiltonian}
\end{equation}
where the $\chi_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\Omega$ couples the two modes and allows for population transfer between them.

The system is prepared with all the population in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to be initially a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number.  As the system is insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, or a mixture of Poissonian-distributed number states.  This state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

@@

In Figure \ref{figTwoModeAnalyticExamples} we plot a solution for the normalized number variance in mode $|a\rangle$ as a function of $\phi$, choosing a typical set of nonlinearities. This example illustrates the basic features of the system: during the final coupling period, the normalised number variance undergoes cyclic variation with a period $\tau=\pi / \Omega$, and has a series of minima. It is necessary to choose both the correct phase $\phi$ as well as the correct length of time for the coupling to be applied in order to obtain optimum squeezing.

@@

The first problem is present even if the spatial modes are uncoupled.  The two-mode model shows that the parameters required to obtain best squeezing are dependent on the strength of the nonlinearity. In the multimode case, the effective nonlinearity of a mode is a function of the mode shape, as is shown by Eq.~(\ref{eqChiUequivalence}). The optimum hold time $\tau_{\mathrm{hold}}$ depends on the product of the effective mode nonlinearity and the number of particles in that mode, as does the recombination phase $\phi$.  Different modes in a multimode environment will typically have varying effective nonlinearity-particle number products and therefore squeeze at different rates, with each mode having a different optimum $\tau_{\mathrm{hold}}$ and optimum recombination phase $\phi$. This means that the best squeezing will be achieved by choosing values of $\tau_{\mathrm{hold}}$ and $\phi$ that, averaged across all the spatial modes present, result in the overall lowest total number variance. Clearly, however, this averaged number variance will be still be higher than the specific mode that exhibits the best squeezing.

@@

An analytic solution to the multimode, higher-dimensional problem is intractable, so we must use a numerical solution.  We use stochastic methods based on phase space representations to simulate the dynamics of the multimode quantum fields.  These methods alleviate the problem of directly simulating elements of the Hilbert space, the size of which increases exponentially with the number of spatial modes \cite{gardiner1991, steelET1998}.  Stochastic methods achieve this by finding a sufficiently well-behaved quasi-probability representation for the density matrix, which can then be simulated as the average behaviour of a number of low-dimensional samples.  These sample objects are the same size as the equivalent classical field.  In our case, the only tractable phase space representation is the functional Wigner representation \cite{steelET1998}, which is well behaved when the quantum field is approximately Gaussian, as it is for coherent and squeezed states.  Stochastic methods based on the functional Wigner representation have been used to model the behaviour of BECs and atom lasers \cite{johnssonET2007,dallET2009,dennisET2010}, as the Wigner representation allows the mapping of the system to a Fokker-Planck equation with positive semi-definite diffusion matrix (unlike the $P$-representation \cite{gardiner1991}), and does not suffer exponential path-weighting problems that would constrain it to very short simulation intervals (as would the positive-$P$ representation \cite{gardiner1991}).

@@

The simulation shown in Fig.~\ref{fig:3DBogScaling} verified that the number of accessible states for three-dimensional traps decreases with volume in 3D.  This is because the energy of the ground state of the BEC is reducing faster than the density of states at low energies is increasing.  This scaling of the estimated Bogoliubov fraction in 3D suggests that strong squeezing may actually be found in very large, weakly trapped condensates, provided the trapping potentials are box-like.  Although systematic analysis of squeezing across a wide range of large 3D traps and interaction strengths is impossible due to the breakdown of the TWA method, fortunately the method is accurate when the highest squeezing is produced.  While inter-state scattering lengths of most atomic species and states are not well characterised, extremely high squeezing can be found in systems with large, three-dimensional box-like potentials, and asymmetry in the three scattering lengths.  A sample 3D simulation with this geometry \textbf{with some lengths and some atom number}, and only a single non-zero scattering length, gives a result of 15dB of squeezing.  This system produces an extremely low non-condensed fraction of $4 \times 10^{-4}$, in agreement with the Bogoliubov calculations.  
%From the "winner winner chicken dinner" email:  Simulating a 3D BEC in a box with the same parameters as the experimental paper works well. Bog mode fraction is around 4e-4, and we get squeezing of around 0.026. This is assuming we can set two scattering lengths to zero, and have the third equal to that of Rb87.
